Delay Comparison of Different Switch Architectures

### 2013-11-22 MSReport/MSThesis.tex
cosmetic updates to my figures and a few important details updated.

\documentclass[11pt]{article}%{IEEEtran}%[conference]{IEEEtran}%\documentclass[10pt]{article}%{article}{report}{letter}{book}{proc}{slides}%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{wasysym}
%\usepackage{skull}

@@

Although delay is clearly a very important metric to consider in these settings, it is not as readily analyzed as throughput, which makes it more difficult to give a reliable a design recommendation.  In order to model the effects of different switch schedulers we have built a network simulator using assumptions meant to mimic a datacenter environment.  We have compared the performance of two relevant scheduling architectures in the hopes of demonstrating that schedulers that are provably throughput optimal have better delay performance when switches are subjected to high loads.

@@

The motivation behind SLIP was to develop a simple protocol that avoids the use of randomization in the scheduler.  SLIP achieves this by adding a very simple state to the switch.  Each input keeps track of the output to which it last transmitted a packet.  Similarly each output keeps track of the last input from which it received a packet.  Armed with this information,the switch follows these three phases in each contention slot:\\

@@

After the last contention slot, the scheduled input and outputs transmit a single packet and update their state.  Once the packets have been successfully transmitted, the cycle continues.  Note that since the time it takes to transmit a packet depends on its length, there will be a lot of idle time in the system if packet length varies significantly.  In practice this is addressed by breaking packets up into small equal sized cells so scheduling can be done on a cell by cell basis as opposed to a packet by packet basis.  In our simulations we consider both the case of all packets being the same size, and the case in which there is a bimodal distribution of packet sizes.  The overhead necessary to split packets into smaller cells involves adding a header to each cell, and using a larger number of contention slots since they will be needed to schedule per cell size, not packet size.  In our simulations we assumed both of these contributed only a negligible amount to the total delay.

@@

Still ideal QCSMA, is guaranteed to be full throughput.  Unfortunately the delay performance of priority flows will depend on the backlog of the queues which they enter, which penalizes low rate flows.  The ideal QCSMA takes place in continuous time: so it can run arbitrarily fast, as well as being immune to collisions, and thus represents a performance bound for the other schedulers.\\

@@

At the beginning of each transmission opportunity, each input $s$ generates an exponential timer $X_{s,d}$ for each nonempty queue  destined for output $d$ (denoted as $Q_{s,d}$).  $X_{s,d}$ is distributed with rate $\lambda_{s,d} = (Q_{s,d})^{\alpha}$.  Where $\alpha > 0$ is a parameter to be chosen.  Input $s$ then requests access to output $d$ when timer $X_{s,d}$ expires. \\ %There will be no collisions  as this happens in continuous time.\\

{\bf Ideal QCSMA:}
\begin{itemize}
\item {\it unscheduled} input queue $Q_{s,d}$ generates timer value $X_{s,d}\sim \text{exp}(Q_{s,d}^\alpha)$
\item When $X_{s,d}$ expires schedule transmission from input $s$ to output $d$ unless $d$ or $s$ is already scheduled.
\item When all timers expire transmit packets.\\
\end{itemize}

@@

If each time slot is assumed to have duration $\beta$, then the probability that an exponential timer of rate $Q_{s,d}^\alpha$ expires within a time slot is given by:
\begin{align}
p'_{s,d} =1-e^{\beta Q_{s,d}^\alpha} \approx \beta Q_{s,d}^\alpha
\end{align}
% in  geometric random variable used to approximate the exponential can approximated by:
If we take $p'_{s,d}$ to be the request probability of the input $s$ to output $d$.  When queues get arbitrarily large, the collisions in the system may get excessively large (and therefore block transmissions indefinitely), so we limit the aggressiveness of the request rate in order to limit the number of collisions by defining:
\begin{align}  \label{geo_p}
p_{s,d} =\min \left[ \beta Q_{s,d}^\alpha,p_{\text{cap}}\right]
\end{align}

@@

{\bf Slotted QCSMA:}
\begin{itemize}
\item In the first contention slot for the packet transmission opportunity each input queue calculates $p_{s,d}$
\item {\it unscheduled} input queue $Q_{s,d}$ requests output $d$ with probability $p_{s,d}$.
\item Output $d$ announces the outcome of requests to it.  If there is only one request, $d$ reports success.  Upon multiple requests, $d$ reports collision. For no requests, $d$ reports no transmission.
\item If the requests of input $s$ were successful to the outputs in the set $\mathcal{D}$, $s$ schedules itself to output $d$ with probability $\frac{Q_{s,d}}{\sum_{d\in \mathcal{D}}Q_{s,d}}$ (i.e. $s$ chooses among its transmission opportunities in proportion to the queue lengths)
\item Unsuccessful inputs update $p_{s,d}$ according to one of the schemes below, and re start requesting process\\
\end{itemize}

@@

To compare the delay performance of the different schedulers we built an event driven simulators in C++.  All the graphs were plotted for $32 \times 32$ switches (i.e. 32 inputs and 32 outputs).  Simulation time is discrete with the minimum time step assumed to be the length of one of the bit transmissions of one of the schedulers.   The packet lengths are given in lengths of these time steps.  Because modern crossbars are transmit packets in parallel we assume that it takes 5 clock ticks to transmit a 64 byte packet, and 120 clock ticks to transmit a 1500 byte packet.  For all the SLIP simulations we assumed there were $6$ contention slots used to generate the crossbar schedules, but that these calculations were pipelined in such a way as to make the net scheduling was negligible compared to the packet transmission time.  Simulation inputs were flow patterns passing through the switch, which were identical for all switches. %The switch operation was simulated for $10^6$ time steps, in order to reach the steady state switch behavior%during a given experiment in order to avoid  make the comparison between different scheduling policies and module types without worrying about inconsistencies in flow statistics.

@@

\begin{figure}%[h]
	 \includegraphics[width=90mm]{us_load.eps}
	\caption{Delay vs throughput graphs for loads in a scenario in which all packet have the same length.  SLIP performs comparably to the ideal QCSMA scheduler, and outperforms the realistic implementation of QCSMA.  The parameters for the QCSMA based schedulers are taken from table \ref{qcsma_parameters}.} 	
	\label{one_size}
\end{figure}

\begin{figure}%[h]
	 \includegraphics[width=90mm]{vs_load.eps}
	 \caption{Delay vs throughput graphs for loads in a scenario in which every packet has a probability of $1/2$ of being the minimum size and $1/2$ of being the maximum size. With the variable sized packets SLIP no longer outperforms the practical version of QCSMA.  QCSMA still performs similarly to the single length scenario.  (Note that the x axis in this graph is different from that in figure \ref{one_size}.) The parameters for the QCSMA based schedulers are taken from table \ref{qcsma_parameters}.} 	
	\label{variable_size}
\end{figure}

@@

As the throughput supported by the different switches seems comparable, we focus on the delay performance of the switches under TCP connections.  TCP Inputs to the simulator can be represented as a flow type matrix such as the one visualized in figure \ref{typical_flows}.  The entry $s,d$ represents the type of flow from input $s$ to output $d$.  We consider two main types of flows, high throughput flows that are relatively insensitive to delay (type 1 flows) and low throughput flows that are sensitive to delays (type 2 flows).  The packets generated for TCP transmission are generated according to Markov ON OFF processes much as in the open loop simulations with one producing a very low throughput set of packets and the other being generating a much higher throughput, in order to simulate the different types of datacenter flows.  The type 1 process has transitions from ON to OFF transition with probability $.3$ and from OFF to ON with probability $.7$.  The type 2 process transitions from ON to OFF with probability $0.0781$ and from OFF to ON with probability $0.9219$.  When the process is ON it generates packets, and it does not when it is OFF.   State transitions occur at each time step.

@@

\subsubsection{Parameter Search}

In our model there is no maximum buffer size, and consequently no dropped packets.  Consequently, we need to give TCP feedback on when to halve its window size by another means.  We do this by using the explicit congestion control bit in the tcp header.  Our simulations mark packets as facing congestion with a probability $p_{\text{ecn}}$ if the packets arrive at a queue at or exceeding a threshold $T_{\text{ecn}}$ packets.  After our TCP receives an ack which announces that it faced congestion, it will halve it's window, and wait one round trip time before rehalving it's window, in order to flush out all the duplicate congestion announcements due to packets sent before the congestion announcement was received.

@@

\begin{figure}%[h]
	 \includegraphics[width=90mm]{all2allflows.png}
	\caption{This graph shows the flows experienced by a core router with all to all traffic the red flows are type 2 while the blue flows are type 1.}
	\label{all_to_all_flows}
\end{figure}

\begin{figure}%[h]
	 \includegraphics[width=\textwidth]{all_2_all.eps}
	\caption{This graph shows the delays experienced by a core router with all to all traffic.}
	\label{all_to_all}
\end{figure}

\begin{figure}%[h]
	 \includegraphics[width=90mm]{spread.png}
	\caption{This graph shows a typical spread aggregate pattern.  In this case the blue flows indicate the absence of a flow..}
	\label{spreadagg}
\end{figure}

\begin{figure}%[h]
	 \includegraphics[width=\textwidth]{vs_spread.eps}
	\caption{This graph shows the delay spread of type 1 flows under a simulated spread and aggregate pattern.  Note the difference in average performance (as represented by the red dot).}
	\label{vs_spreadagg}
\end{figure}

\begin{figure}%[h]
	 \includegraphics[width=90mm]{typ_vs_flows.png}
	\caption{This graph represents the flow matrix of a `typical' datacenter load.  Where $80\%$ of the flows (type 2) generate a low volume of traffic and are sensitive to delays.  The other $20\%$ of flows (type 1) generate a large volume of traffic and are require a high throughput but not necessarily low delays.  The color in the $s,d$th position signifies the type of flow going from input $s$ to output $d$.  Flows of type $0$ do not generate any packets.} 	
	\label{typical_flows}
\end{figure}

\begin{figure}%[h]
	 \includegraphics[width=\textwidth]{typ_vs.eps}
	\caption{This graph shows the delays experienced by a typical set of datacenter flows.}
	\label{typ_vs}
\end{figure}

\begin{figure}%[h]
	 \includegraphics[width=\textwidth]{typ_spread_vs.eps}
	\caption{This graph shows the delays experienced by a typical set of datacenter flows with variable packet sizes and two spread aggregate jobs.}
	\label{typ_spread_vs}
\end{figure}

@@

In our simulations for this scenario (see figure \ref{typ_vs}) QCSMA on average performs better than SLIP as usual, but also in the extremes.  The overall delays are also much less than in the other scenarios (an order of magnitude less in fact).  This is most likely due to the lower rate of competition between flows in this scenario.  This is one of the best flow patterns for both the time slotted QCSMA, which can be seen by the fact that it almost matches the performs of it's unimplementable ideal version.

@@

As was to be expected, SLIP once again performed slightly worse, but not significantly.
\begin{figure}%[h]
	 \includegraphics[width=90mm]{typ_spread_flows.png}
	\caption{This graph represents the mix of a `typical' load and two spread and aggregate jobs.}
	\label{typ_spread_flows}
\end{figure}

@@

\label{bib}
\bibliography{MSThesis}{}
\bibliographystyle{plain} 
 \newpage
\appendix
%  \appendices 
 \section{Simulation Design} \label{code_design}

>>>>

\documentclass[11pt]{article}%{IEEEtran}%[conference]{IEEEtran}%\documentclass[10pt]{article}%{article}{report}{letter}{book}{proc}{slides}%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{wrapfig}

%\usepackage{skull}

@@

Although delay is clearly a very important metric to consider in these settings, it is not as readily analyzed as throughput, which makes it more difficult to give a reliable design recommendation.  In order to model the effects of different switch schedulers we have built a network simulator using assumptions meant to mimic a datacenter environment.  We have compared the performance of two relevant scheduling architectures in the hopes of demonstrating that schedulers that are provably throughput optimal have better delay performance when switches are subjected to high loads.

@@

The motivation behind SLIP was to develop a simple protocol that avoids the use of randomization in the scheduler.  SLIP achieves this by adding a very simple state to the switch.  Each input keeps track of the output to which it last transmitted a packet.  Similarly each output keeps track of the last input from which it received a packet.  Armed with this information, the switch follows these three phases in each contention slot:\\

@@

After the last contention slot, the scheduled inputs and outputs transmit a single packet and update their state.  Once the packets have been successfully transmitted, the cycle continues.  Note that since the time it takes to transmit a packet depends on its length, there will be a lot of idle time in the system if packet length varies significantly.  In practice this is addressed by breaking packets up into small equal sized cells so scheduling can be done on a cell by cell basis as opposed to a packet by packet basis.  In our simulations we consider both the case of all packets being the same size, and the case in which there is a bimodal distribution of packet sizes.  The overhead necessary to split packets into smaller cells involves adding a header to each cell, and using a larger number of contention slots since they will be needed to schedule per cell size, not packet size.  In our simulations we assumed both of these contributed only a negligible amount to the total delay.

@@

Ideal QCSMA is guaranteed to be full throughput.  Unfortunately the delay performance of flows will depend on the backlog of the queues which they enter, which penalizes low rate flows.  The ideal QCSMA takes place in continuous time: so it can run arbitrarily fast, as well as being immune to collisions, and thus represents a performance bound for the other schedulers.\\

@@

At the beginning of each transmission opportunity, each input $s$ generates an exponential timer $X_{s,d}$ for each nonempty queue  destined for output $d$ (denoted as $Q_{s,d}$).  $X_{s,d}$ is exponentially distributed with rate $\lambda_{s,d} = (Q_{s,d})^{\alpha}$.  Where $\alpha > 0$ is a parameter to be chosen.  Input $s$ then requests access to output $d$ when timer $X_{s,d}$ expires. \\ %There will be no collisions  as this happens in continuous time.\\

{\bf Ideal QCSMA:}
\begin{itemize}
\item {\it unscheduled} input queue $Q_{s,d}$ generates timer value $X_{s,d}\sim \text{exp}(Q_{s,d}^\alpha)$
\item When $X_{s,d}$ expires schedule transmission from input $s$ to output $d$ unless $d$ or $s$ is already scheduled.
\item When a timer expires transmit packets.\\
\end{itemize}

@@

If each time slot is assumed to have duration $\beta$, then the probability that an exponential timer of rate $Q_{s,d}^\alpha$ expires within a time slot is given by:
\begin{align}
p'_{s,d} =1-e^{\beta Q_{s,d}^\alpha} \approx \beta Q_{s,d}^\alpha
\end{align}
% in  geometric random variable used to approximate the exponential can approximated by:
We take $p'_{s,d}$ to be the request probability of the input $s$ to output $d$.  When queues get arbitrarily large, the collisions in the system may get excessively frequent (and therefore block transmissions indefinitely), so we limit the aggressiveness of the request rate in order to limit the number of collisions by defining:
\begin{align}  \label{geo_p}
p_{s,d} =\min \left[ \beta Q_{s,d}^\alpha,p_{\text{cap}}\right]
\end{align}

@@

{\bf Slotted QCSMA:}
\begin{itemize}
\item In the first contention slot for the packet transmission opportunity each input queue calculates $p_{s,d}$
\item {\it unscheduled} input queue $Q_{s,d}$ requests output $d$ with probability $p_{s,d}$.
\item Output $d$ announces the outcome of requests to it.  If there is only one request, $d$ reports success to that input, otherwise $d$ reports collision.%  Upon multiple requests, $d$ reports collision. For no requests, $d$ reports no transmission.
\item If the requests of input $s$ were successful to the outputs in the set $\mathcal{D}$, $s$ schedules itself to output $d$ with probability $\frac{Q_{s,d}}{\sum_{d\in \mathcal{D}}Q_{s,d}}$ (i.e. $s$ chooses among its transmission opportunities in proportion to the queue lengths)
\item Unsuccessful inputs could update $p_{s,d}$ according to one of the schemes below, and re start requesting process\\
\end{itemize}

@@

To compare the delay performance of the different schedulers we built an event driven simulators in C++.  All the graphs were plotted for $32 \times 32$ switches (i.e. 32 inputs and 32 outputs).  Simulation time is discrete with the minimum time step assumed to be the length of one of the bit transmissions of one of the schedulers.   The packet lengths are given in lengths of these time steps.  Because modern crossbars transmit packets in parallel we assume that it takes 5 clock ticks to transmit a 64 byte packet, and 120 clock ticks to transmit a 1500 byte packet.  For all the SLIP simulations we assumed there were $6$ contention slots used to generate the crossbar schedules, but that these calculations were pipelined in such a way as to make the net scheduling negligible compared to the packet transmission time.  Simulation inputs were flow patterns passing through the switch, which were identical for all switches. %The switch operation was simulated for $10^6$ time steps, in order to reach the steady state switch behavior%during a given experiment in order to avoid  make the comparison between different scheduling policies and module types without worrying about inconsistencies in flow statistics.

@@

\begin{figure}
\center
	 \includegraphics[width=90mm]{us_load.eps}
	\caption{Delay vs throughput graphs for loads in a scenario in which all packet have the same length.  SLIP performs comparably to the ideal QCSMA scheduler, and outperforms the realistic implementation of QCSMA.  The parameters for the QCSMA based schedulers are taken from table \ref{qcsma_parameters}.} 	
	\label{one_size}
\end{figure}

\begin{figure}
\center
	 \includegraphics[width=90mm]{vs_load.eps}
	 \caption{Delay vs throughput graphs for loads in a scenario in which every packet has a probability of $1/2$ of being the minimum size and $1/2$ of being the maximum size. With the variable sized packets SLIP no longer outperforms the practical version of QCSMA.  QCSMA still performs similarly to the single length scenario.  (Note that the x axis in this graph is different from that in figure \ref{one_size}.) The parameters for the QCSMA based schedulers are taken from table \ref{qcsma_parameters}.} 	
	\label{variable_size}
\end{figure}

@@

As the throughput supported by the different switches seems comparable, we focus on the delay performance of the switches under TCP connections.  TCP Inputs to the simulator can be represented as a flow type matrix such as the one visualized in figure \ref{typical_flows}.  The entry $s,d$ represents the type of flow from input $s$ to output $d$.  We consider two main types of flows, high throughput flows that are relatively insensitive to delay (type 1 flows) and low throughput flows that are sensitive to delays (type 2 flows).  The packets generated for TCP are generated according to Markov ON OFF processes much as in the open loop simulations with one producing a very low throughput set of packets and the other being generating a much higher throughput, in order to simulate the different types of datacenter flows.  The type 1 process has transitions from ON to OFF transition with probability $.3$ and from OFF to ON with probability $.7$.  The type 2 process transitions from ON to OFF with probability $0.0781$ and from OFF to ON with probability $0.9219$.  When the process is ON it generates packets, and it does not when it is OFF.   State transitions occur at each time step.

@@

\subsection{Parameter Search}

%In our model there is no maximum buffer size, and consequently no dropped packets.  Consequently, we need to give TCP feedback on when to halve its window size by another means.  We do this by using the explicit congestion control bit in the tcp header.  Our simulations mark packets as facing congestion with a probability $p_{\text{ecn}}$ if the packets arrive at a queue at or exceeding a threshold $T_{\text{ecn}}$ packets.  After our TCP receives an ack which announces that it faced congestion, it will halve it's window, and wait one round trip time before rehalving it's window, in order to flush out all the duplicate congestion announcements due to packets sent before the congestion announcement was received.

@@

\begin{figure}
\center
	 \includegraphics[width=90mm]{all2allflows.png}
	\caption{This graph shows the flows experienced by a core router with all to all traffic the red flows are type 2 while the blue flows are type 1.}
	\label{all_to_all_flows}
\end{figure}

\begin{figure}
\center
	 \includegraphics[width=\textwidth]{all_2_all.eps}
	\caption{This graph shows the delays experienced by a core router with all to all traffic.}
	\label{all_to_all}
\end{figure}

\begin{figure}
\center
	 \includegraphics[width=90mm]{spread.png}
	\caption{This graph shows a typical spread aggregate pattern.  In this case the blue flows indicate the absence of a flow..}
	\label{spreadagg}
\end{figure}

\begin{figure}
\center
	 \includegraphics[width=\textwidth]{vs_spread.eps}
	\caption{This graph shows the delay spread of type 1 flows under a simulated spread and aggregate pattern.  Note the difference in average performance (as represented by the red dot).}
	\label{vs_spreadagg}
\end{figure}

\begin{figure}
\center
	 \includegraphics[width=90mm]{typ_vs_flows.png}
	\caption{This graph represents the flow matrix of a `typical' datacenter load.  Where $80\%$ of the flows (type 2) generate a low volume of traffic and are sensitive to delays.  The other $20\%$ of flows (type 1) generate a large volume of traffic and are require a high throughput but not necessarily low delays.  The color in the $s,d$th position signifies the type of flow going from input $s$ to output $d$.  Flows of type $0$ do not generate any packets.} 	
	\label{typical_flows}
\end{figure}

\begin{figure}
\center
	 \includegraphics[width=\textwidth]{typ_vs.eps}
	\caption{This graph shows the delays experienced by a typical set of datacenter flows.}
	\label{typ_vs}
\end{figure}

\begin{figure}
\center
	 \includegraphics[width=\textwidth]{typ_spread_vs.eps}
	\caption{This graph shows the delays experienced by a typical set of datacenter flows with variable packet sizes and two spread aggregate jobs.}
	\label{typ_spread_vs}
\end{figure}

@@

In our simulations for this scenario (see figure \ref{typ_vs}) QCSMA on average performs better than SLIP as usual, but also in the extremes.  The overall delays are also much less than in the other scenarios (an order of magnitude less in fact).  This is most likely due to the lower rate of competition between flows in this scenario.  This is one of the best flow patterns for both the time slotted QCSMA, which can be seen by the fact that it almost matches the performance of it's unimplementable ideal version.

@@

As was to be expected, SLIP once again performed slightly worse, but not significantly.
\begin{figure}%[h]
\center
	 \includegraphics[width=90mm]{typ_spread_flows.png}
	\caption{This graph represents the mix of a `typical' load and two spread and aggregate jobs.}
	\label{typ_spread_flows}
\end{figure}

@@

\label{bib}
\bibliography{MSThesis}{}
\bibliographystyle{plain} 
 \newpage
\appendix
%  \appendices 
 \section{Simulator Design} \label{code_design}
